Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-10-11 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)
library(mice)         # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel pooling
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

# Case study II: HIV data

[TODO: change the order of case study 2 and 3]

```{r hiv, echo=FALSE}
# data
data("hiv", package = "GJRM")
hiv <- hiv %>% 
  select(c("region","hiv", "age", "marital", "condom", "smoke")) %>% 
  mutate(region = as.integer(region))
# md pattern
plot_pattern(hiv)
```

# Case study III: IMPACT data

@@

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification,
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

@@

## Complete case analysis

@@

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor:
```{r ignore}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred)
```

@@

```{r imp_impact, eval = FALSE, include=FALSE}
imp <- mice(dat, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
```

```{r impact-fit, eval = FALSE, include=FALSE}
fit <- imp_impact_pmm %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

>>>>

[TODO: Add environment info, seed and version number(s) somewhere.] Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)
library(mice)         # for imputation
library(miceadds)     # for imputation
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel pooling
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

# Case study II: IMPACT data (syst missingness, pred matrix)

@@

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification, [TODO: make this one var? also shows that you don't always need random effects everywhere?]
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

@@

## Complete case analysis [TODO: remove this?]

@@

Create a methods vector and predictor matrix, and make sure `name` is not included as predictor, but as clustering variable:
```{r ignore}
meth <- make.method(dat) # methods vector
pred <- quickpred(dat)   # predictor matrix
plot_pred(pred)
```

@@

```{r imp_impact, eval = FALSE, include=FALSE}
dat <- dat %>% mutate(across(everything(), as.integer))
imp <- mice(dat, method = meth, predictorMatrix = pred, m = 2, maxit = 1)
```

```{r impact-fit, eval = FALSE, include=FALSE}
fit <- imp %>% 
  with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
tidy(pool(fit))
as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```

# Case study III: HIV data

```{r hiv, echo=FALSE}
# data
data("hiv", package = "GJRM")
hiv <- hiv %>% 
  select(c("region","hiv", "age", "marital", "condom", "smoke", "interviewerID")) %>% 
  mutate(region = as.integer(region))
# md pattern
plot_pattern(hiv)
```

R=region+language+(1|InterviewID) model with with interviewer as random effects, because the observations are not independent. Interviews are not allocated randomly. In theory we expect the inclusion-restriction variable to be randomly assigned, that's why we're adding region and language to compensate for non-random allocation.

```{r eval=FALSE, include=FALSE}
meth <- make.method(hiv)
meth["hiv"] <- "2l.heckman"
imp <- mice(hiv, method = meth)
pred <- make.predictorMatrix(hiv)
pred[, "region"] <- -2
pred[, "smoke"] <- -3
imp2 <- mice(hiv, method = meth, predictorMatrix = pred)
```

```{r}
data("hiv",  package = "GJRM")
# We select  5 predictor variables over 9 regions
colnames(hiv)
hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
# Study/group variable has to be recoded as integer
hivdata$region<-as.integer(hivdata$region)

# Categorical variables have to be recoded as factor
# to use the 2l.binary imputation method, it is required that the level names should not have any space between words

hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
#the missing variable is stored as a factor in the dataset, otherwise the method will 
#apply the imputation correction for a  missing continuous variable instead for 
#a missing binary variable,which is in this case the binary response of the hiv test.

hivdata$marital <- as.factor(hivdata$marital)
levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
hivdata$condom <- as.factor(hivdata$condom)
levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
hivdata$highhiv <- as.factor(hivdata$highhiv)
levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
hivdata$interviewerID <- as.factor(hivdata$interviewerID)

hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))

summary(hivdata)
#  We obtain here the random effects for each interviewer, this is an approximation 
#of the interviewer's skill which will be used as an exclusion constraint. 
#Here, since the location of the interviewer was not randomly assigned to the 
#subjects, the assignment was corrected for region and language.

hivdata$hivconsent <- as.factor(hivdata$hivconsent)
ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
reffect  <- ranef(ID_mixed)$interviewerID
reffect$interviewerID <- levels(hivdata$interviewerID)
colnames(reffect)     <- c("IDreffect","interviewerID")
hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)

hivdata$interviewerID<-NULL
hivdata$hivconsent<-NULL
hivdata$language<-NULL
summary(hivdata)

# #Set prediction matrix and methods
# ini <- mice(hivdata, maxit = 0)
# meth<-ini$method
# meth["hiv"]<-"2l.binary"
# pred <- ini$pred
# pred[,"region"] <- 0
# pred["region",] <- 0
# pred["hiv","region"]<- -2
# pred["hiv","IDreffect"]<- 0
# 
# # MAR imputation model, we used the 2l.binary from the miceadds package
# data_imp_mar <- mice( data= hivdata, # dataset with missing values
#                       m = 10,   # number of imputations
#                       meth = meth, #imputation method vector
#                       pred = pred, #imputation predictors matrix
#                       maxit=1)
# 
# # source("mice.impute.2l.heckman.R")
# # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["1"]]))
# # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["1"]]))
# # sum(as.numeric(data_imp_mar[["imp"]][["hiv"]][["2"]]))
# # sum(as.numeric(data_imp_heckp[["imp"]][["hiv"]][["2"]]))
# 
# # Heckman model
# pred["hiv","IDreffect"] <- -3
# meth<-ini$method
# meth["hiv"]<-"2l.heckman"
# 
# data_imp_heckp <- mice( hivdata, # dataset with missing values
#                         m = 10,   # number of imputations
#                         seed = 1234, #seed attached to the dataID
#                         meth = meth, #imputation method vector
#                         pred = pred, #imputation predictors matrix
#                         print = T,
#                         meta_method="reml",
#                         pmm=FALSE)

#Set prediction matrix and methods
ini <- mice(hivdata, maxit = 0)
meth<-ini$method
meth["hiv"]<-"2l.binary"
pred <- ini$pred
pred[,"region"] <- 0
pred["region",] <- 0
pred["hiv","region"]<- -2
pred["hiv","IDreffect"]<- 0

# MAR imputation model, we used the 2l.binary from the miceadds package
data_imp_mar <- mice( data= hivdata, # dataset with missing values
                      m = 20,   # number of imputations
                      meth = meth, #imputation method vector
                      pred = pred, #imputation predictors matrix
                      maxit=1)

source("mice.impute.2l.heckman.R")

# Heckman model
pred["hiv","IDreffect"] <- -3
meth<-ini$method
meth["hiv"]<-"2l.heckman"

data_imp_heckp <- mice( hivdata, # dataset with missing values
                        m = 20,   # number of imputations
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        maxit=1,
                        meta_method="reml",
                        pmm=FALSE)
```
